Whole-Body Manipulation Planning for Humanoid Robots

### 2012-06-15 paper.tex
Bibliography correction

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Momentum
Point (ZMP) formalism. This section summarizes the related work for each of them.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Simplifying  the environment  consists in  considering obstacles  at a
footstep   level only.   \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
use   an  A$^{*}$   algorithm  to   find
collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
collision-free motions  for the legs by using  an RRT$^{*}$ algorithm.

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not directly applicable to
humanoid dynamically stable locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. In our work, we show that it is not always
necessary to take into account dynamics while planning, even though the final
output of our algorithm is a dynamically stable walk motion.

@@

The idea of this proof is to use the form of Eq. (\ref{eq:walk-zmp}) to
apply a  scaling factor between  the amplitude of the  oscillations of
the CoM and of the ZMP. For $\omega >0$, assuming the CoM follows the trajectory
$y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
\[
p_y(t) =
(1+\left(\frac{\omega}{\omega_0}\right)^2)\epsilon\sin(\omega t)
\]

@@

\paragraph{Use of ZMP preview controller:} The control strategy presented in the  
previous proof may
generate very  long trajectories, because  of the transient  states at
the beginning and end of the locomotion. In the actual implementation,
we have chosen  to generate  CoM motions with  a ZMP preview  controller, as
presented in \cite{kajita2003biped}.  We  have observed experimentally that the
amplitude of  CoM trajectories decreases  when the frequency  of steps
increases. Our current ZMP preview controller relies on the cart-table simplified model.
To make this approximation valid, we fix the height of the robot CoM during walk,
as well as the vertical orientation of the robot waist. These geometric constraints 
are also applied when planning a statically stable path. Note that this is
due to our current implementation of ZMP preview controller, and does not affect
the generality of the small-space controllability result presented above.

@@

The  algorithm   that  animates  a  statically  stable   path  into  a
dynamically stable  walk trajectory has been inspired  by the previous
small-space controllability  proof.
Given a statically  stable path $p$,
we  start  by placing  footsteps  corresponding  to  the nominal  walk
pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
and a preview controller outputs  a corresponding CoM trajectory, using
the cart-table model. The stack of tasks applied to
the robot to generate a dynamic walk motion is -- in decreasing priority order:

@@

\item Upper-body configuration task towards corresponding
  configuration  of $p$.

@@

The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} was
presented in \cite{el2011path}. The  authors solved it by using a bounding
box method, leading the robot to walk sideways between the two chairs.
Our method generated a locomotion  trajectory in which the robot walks
forward, which could be required if the robot has to use vision during
locomotion for  example. The first planning stage  required 29.6~s
on  average.   The  animation  of   the  sliding  path   presented  in
Fig.   \ref{fig:sliding}    used   66.5~s  of  computation   time.

@@

In the environment shown in Fig. \ref{fig:cluttered}, the robot
has to find a way among floating obstacles. In this
environment neither bounding box nor footstep planning strategies
could find a collision-free walk trajectory.
The first planning stage required
184.3~s on average, and the animation of the trajectory presented in 
Fig.~\ref{fig:cluttered} used 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
shows the robot CoM trajectory during locomotion.

@@

In the problem shown in Fig. \ref{fig:shelf} the robot has to put a ball on a
shelf, in a constrained apartment environment. The final configuration is defined 
implicitly as a desired hand position. We generated automatically goal configurations 
solving the task, as described in Section~\ref{sec:wb} . Then, we 
applied our planner to generate a whole-body walk motion that solved the hand reaching
task.

>>>>

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Momentum
Point (ZMP) formalism. This section summarizes the literature related to each of 
these fields.

@@

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
overviews.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

Simplifying  the environment  consists in  considering obstacles  at a
footstep   level only.   \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
use   an  A$^{*}$   algorithm  to   find
collision-free   footsteps.    In   \cite{perrin2012fast},   the   authors   compute
dynamic walking motions avoiding collisions at the leg level
by using  an RRT algorithm.

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not directly applicable to
humanoid dynamically stable locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. To the authors knowledge such techniques have not been
used on humanoid robotic platforms so far.
In our work, we show that it is not always
necessary to take into account dynamics while planning, even though the final
output of our algorithm is a dynamically stable walk motion. This result was first
presented in \cite{dalibard2011small}.

@@

The idea of this proof is to use the form of Eq. (\ref{eq:walk-zmp}) to
apply a  scaling factor between  the amplitude of the  oscillations of
the CoM and of the ZMP. The faster the CoM oscillates, the bigger is the amplitude
of the ZMP oscillations. Following is a formalization of this  idea.

For $\omega >0$, assuming the CoM follows the trajectory
$y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
\[
p_y(t) =
(1+\left(\frac{\omega}{\omega_0}\right)^2)\epsilon\sin(\omega t)
\]

@@

\paragraph{Use of ZMP preview controller:} The control strategy presented in the  
previous proof may
generate very  long trajectories, because  of the transient  states at
the beginning and end of the locomotion. In the actual implementation,
we have chosen  to generate  CoM motions with  a ZMP preview  controller, as
presented in \cite{kajita2003biped}.  We  have observed experimentally that the
amplitude of  CoM trajectories decreases  when the frequency  of steps
increases. Our current ZMP preview controller relies on the cart-table model approximation.
To make this approximation valid, we fix the height of the robot CoM during walk,
as well as the vertical orientation of the robot waist. These geometric constraints 
are also applied when planning statically stable paths, to ensure that the paths
can be approximated by dynamic walk trajectories. Note that this is
due to our current  ZMP preview controller implementation, and does not affect
the generality of the small-space controllability result presented above.

@@

The  algorithm   that  animates  a  statically  stable   path  into  a
dynamically stable  walk trajectory has been inspired  by the previous
small-space controllability  proof.
Given a statically  stable path $p$ respecting the cart-table model approximation constraints,
we  start  by placing  footsteps  corresponding  to  the nominal  walk
pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
and a preview controller outputs  a corresponding CoM trajectory. 
The stack of tasks applied to
the robot to generate a dynamic walk motion is -- in decreasing priority order:

@@

\item Configuration task towards corresponding
  configuration  in $p$.

@@

The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} has been
first presented in \cite{el2011path}. It was solved by using a bounding
box method, leading the robot to walk sideways between the two chairs.
Our method generates a locomotion  trajectory in which the robot walks
forward, which may be required if the robot has to use vision during
locomotion. The first planning stage  requires 29.6~s
on  average.   The  animation  of   the  sliding  path   presented  in
Fig.   \ref{fig:sliding}   uses  66.5~s  of  computation   time.

@@

In the environment shown in Fig. \ref{fig:cluttered}, the robot
has to find a way among floating obstacles. In this
environment neither bounding box nor footstep planning strategies
could find a collision-free walk trajectory.
The first planning stage requires
184.3~s on average, and the animation of the trajectory presented in 
Fig.~\ref{fig:cluttered} uses 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
shows the robot CoM trajectory during locomotion.

@@

In the problem shown in Fig. \ref{fig:shelf} the robot has to put a ball on a
shelf, in a constrained apartment environment. The final configuration is defined 
implicitly as a desired hand position. We have generated automatically goal configurations 
solving the task, as described in Section~\ref{sec:wb} . Then, we have
applied our planner to generate a whole-body walk motion that solves the hand reaching
task.
